# Cache credentials for HTTPS {#credential-caching}

If you plan to push/pull using HTTPS, you want to cache your credentials (username, password), so you don't need to enter them over and over again.
Alternatively, you could set up SSH keys (chapter \@ref(ssh-keys)).
I suggest you set up one of these methods of authentication on each computer you want to connect to GitHub from.

I find HTTPS easier to get working quickly and **strongly recommend** it when you first start working with Git/GitHub.
[HTTPS is what GitHub recommends](https://stackoverflow.com/a/11041782/2825349), presumably for exactly the same reasons.
I started with HTTPS, preferred SSH for a while, and have returned to HTTPS.
Either is fine, you can change your mind later, and you can use HTTPS on one machine and SSH on another.

## You should get a personal access token (PAT)

### Why a PAT?

Password-based authentication for Git is deprecated, i.e. you really should _not_ be sending your username and password every time you push or pull.
Here, we're talking about the username and password you would use to login to GitHub in the browser.

What should you do instead?

Get a **personal access token** (PAT) and use that as your credential for HTTPS operations.
(The PAT will actually be sent as the password and the username is somewhat artificial, consulted only for credential lookup.)

If you turn on [two-factor authentication](https://docs.github.com/en/free-pro-team@latest/github/authenticating-to-github/about-two-factor-authentication) (a.k.a. "2FA") for GitHub and you use HTTPS, you absolutely **must** send a personal access token.
And, really, it's a good idea for everyone to turn on 2FA and for everyone who uses HTTPS to use a PAT.

The final selling point is that once you configure a PAT, several R packages, including usethis and gh, will be able to work with the GitHub API on your behalf, automagically.

### How to get a PAT?

GitHub offers instructions for [creating a personal access token](https://docs.github.com/en/free-pro-team@latest/github/authenticating-to-github/creating-a-personal-access-token).

The usethis package has a helper function that takes you to the web form to create a PAT, with the added benefit that it pre-selects the recommended scopes:

```{r eval = FALSE}
usethis::create_github_token()
```

Once you are happy with the selected scopes, click "Generate token".
As the page says, you must **store this token somewhere**, because you'll never be able to see it again, once you leave that page or close the window.

### How to manage a PAT?

Treat this PAT like a password!
If you use a password management app, such as 1Password or LastPass (which you should), it is highly recommended to add this PAT to your entry for GitHub.
Below, we will add your PAT to the Git credential store as a semi-persistent convenience, sort of like "remember me" on a website.
But, just like logging into websites, it is entirely possible that your PAT will somehow be forgotten by the credential store and you will need to re-enter it.

If you goof this up, i.e. generate a PAT but fail to capture it on your system,
you'll have to generate another one.
This is not the end of the world, but you should delete the "lost" PAT on GitHub.
If you aren’t disciplined about labelling PATs and deleting lost PATs, you will find yourself in an unsettling situation where you can’t be sure which PAT(s) are in use.

Do not ever hard-wire your PAT into your code!
It should always be retrieved implicitly, for example, from the Git credential store or from an environment variable.

At this point, I assume you've generated a PAT and have it available, either:

  * In a secure, long-term system for storing secrets, like 1Password or LastPass (recommended)
  * For the next few minutes, e.g. in a browser window or on the clipboard

## Store your credentials

There are many ways to get your credentials into the Git store.
You need to trigger a prompt for your credentials and usually they'll be stored for next time.
(If they don't seem to be stored and re-discovered, see the next section.)

Ways to get that initial prompt:

  * Call an R function to store your credentials
  * Organic Git use, e.g. via the command line or even RStudio  
  * Make an explicit call to the Git credential store (an unusual choice, which
  we won't describe)

Pick one!

### Call an R function to store your credentials

As of October 2020, there are two R packages for accessing the Git credential store from R:

  * [gitcreds](https://r-lib.github.io/gitcreds/) *Not yet on CRAN*
  * [credentials](https://docs.ropensci.org/credentials/)
    
It is likely that these packages will eventually combine into one and, even now, they are largely interoperable.
You don't need to follow the instructions for both packages -- pick one!

#### gitcreds

*As of October 2020, gitcreds is not yet on CRAN, but we anticipate a release soon.
usethis 2.0.0 will depend on gitcreds, via the [gh package](https://gh.r-lib.org).*

If you don't have gitcreds installed, install via `devtools::install_github("r-lib/gitcreds")` or, eventually, via `install.packages("gitcreds")`.

Then call `gitcreds_set()`:

```{r eval = FALSE}
library(gitcreds)

gitcreds_set()
```

This should allow you to keep or view any existing credentials and replace them or store a credential for the first time.

Respond to the prompt with your personal access token (PAT).

You can check that you've stored credentials like so:

```{r eval = FALSE}
gitcreds_get()
#> <gitcreds>
#>   protocol: https
#>   host    : github.com
#>   username: PersonalAccessToken
#>   password: <-- hidden -->
```

#### credentials

If you don't have credentials installed, install via `install.packages("credentials")`.

Then call `set_github_pat()`:

```{r eval = FALSE}
library(credentials)

set_github_pat()
```

If successful, your initial (and subsequent) calls will look like this:

```{r eval = FALSE}
set_github_pat()
#> If prompted for GitHub credentials, enter your PAT in the password field
#> Using GITHUB_PAT from Jennifer (Jenny) Bryan (credential helper: osxkeychain)
```

### Store credentials through organic Git use

*Before gitcreds and credentials existed (see above), we had to store the PAT through organic Git use.
We still show this method, but we now recommend using the approaches above, because they are much faster.*

Pre-requisite: You need a functioning test Git repository.
One that exists locally and remotely on GitHub, with the local repo tracking the remote.

If you have just verified that you can interact with GitHub (chapter \@ref(push-pull-github)) from your local computer, that test repo will be perfect.

If you have just verified that you can work with GitHub from RStudio (chapter \@ref(rstudio-git-github)), that test repo will also be perfect.

You may proceed when

  * You have a test repo.
  * You know where it lives on your local computer. Example:
    - `/home/jenny/tmp/myrepo`
  * You know where it lives on GitHub. Example:
    - `https://github.com/jennybc/myrepo`
  * You know the GitHub repo is setup as a remote. In a shell (Appendix \@ref(shell)) working directory set to the local Git repo, enter:
  
        git remote -v
        
    Output like this confirms that fetch and push are set to remote URLs that point to your GitHub repo:
    
        origin	https://github.com/jennybc/myrepo (fetch)
        origin	https://github.com/jennybc/myrepo (push)
        
    Now enter:
    
        git branch -vv
        
    Here we confirm that the local `master` branch is tracking your GitHub master branch (`origin/master`). Gibberish? Just check that your output looks similar to mine:
    
        master b8e03e3 [origin/master] line added locally

Trigger a username / password challenge

Change a file in your local repo and commit it.
Do that however you wish. Here are shell commands that will work:

    echo "adding a line" >> README.md
    git add -A
    git commit -m "A commit from my local computer"

Now push!

    git push -u origin master

You should be asked for your username and password.
If you've taken our advice to get a PAT, **provide your PAT as the password**.
Hopefully, this credential will be stored.

Now push AGAIN.

    git push
  
You should NOT be asked for your username and password, instead you should see `Everything up-to-date`.
  
Rejoice and close the shell.

## Activating a Git credential manager

### You might not need to do anything!

As of October 2020, if you install Git using the [methods recommended here in Happy Git](#install-git), it is likely that Git is already configured to use a credential helper provided by your operating system.
Of course, you will have to provide your credentials at least once, but most users do not need to do anything special to arrange for those credentials to be stored and retrieved later.

Specifically, if you are on macOS or Windows, don't do anything described here until you have actual proof that it's necessary, i.e. until you have experienced repeated challenges for your credentials when using HTTPS.

### Windows

In my most recent experiments, Git for Windows is configured to store credentials via the Windows Credentials Manager.

Here’s a command to reveal the configuration and the output I see in a fresh installation of Git for Windows:

``` bash
$ git config --show-origin --get credential.helper
file:C:/Program Files/Git/mingw64/etc/gitconfig manager
```

In the unlikely event that you need to activate a credential helper, GitHub's instructions are the best bet, in terms of being current:

[Caching your GitHub credentials in Git](https://docs.github.com/en/free-pro-team@latest/github/using-git/caching-your-github-credentials-in-git)

### macOS

I have not needed to explicitly activate credential caching on macOS in a long while.

Here’s a command to reveal the current credential helper and the output I see (October 2020, macOS 1-.15.6, Git 2.24.3):

``` bash
$ git config --show-origin --get credential.helper
file:/Users/jenny/.gitconfig    osxkeychain
```

In the unlikely event that you need to activate a credential helper, GitHub's instructions are the best bet, in terms of being current:

[Caching your GitHub credentials in Git](https://docs.github.com/en/free-pro-team@latest/github/using-git/caching-your-github-credentials-in-git)


### Linux

As opposed to Windows and macOS, Linux users are the ones for whom credential storage may not "just work".
The easiest thing to do is to configure Git to "cache" your credentials (vs "store"), which is more time-limited.
Then set the cache timeout to some suitably long period of time.
You can expect to re-enter your PAT more often than Windows and macOS users.

We paraphrase the Linux instructions from GitHub's documentation [Caching your GitHub credentials in Git](https://docs.github.com/en/free-pro-team@latest/github/using-git/caching-your-github-credentials-in-git), which should be regarded the best and most current source.

In the shell, turn on the "cache" credential helper and set its timeout:

``` bash
git config --global credential.helper 'cache --timeout=10000000'
```

Above, we set the timeout to ten million seconds or around 16 weeks, enough for a semester.
