# Existing project, GitHub last {#existing-github-last}

This an explicit workflow for connecting an existing local R project to GitHub, when for some reason you cannot or don't want to do a "GitHub first" workflow (see chapters \@ref(new-github-first) and \@ref(existing-github-first)).

When does this come up?
Example: it's an existing project that is already a Git repo with a history you care about.
Then you have to do this properly.

This may be less desirable for a novice because there are more opportunities to get confused and make a mistake.
But this workflow is not that hard, even with command line Git, and is even easier if you use conveniences from the [usethis](https://cran.r-project.org/package=usethis) package or the RStudio IDE.
All of these are covered below.

## Prepare the local project

We assume you've got your existing R project isolated in a directory on your computer.
If that's not already true, make it so.
Create a directory and marshal all the existing data and R scripts there.
It doesn't really matter where you do this, but note where the project currently lives.

I encourage you to make this project into an RStudio project, although it is not absolutely required.
If you opt-out of this, the instructions using command line Git or usethis will still work for you, outside of RStudio.

### Make or verify an RStudio Project

If the project is not already an RStudio Project, make it so:

  * Within RStudio you can do: *File > New Project > Existing Directory* and, if you wish, "Open in new session".
  * Alternatively, from R, call `usethis::create_project("path/to/your/project")`, substituting the path to your existing project directory.

If your project is already an RStudio Project, launch it.

### Make or verify a Git repo

You should be in RStudio now, in your project.

Is it already a Git repository?
The presence of the Git pane should tip you off.
If yes, you're done.

If not, you have several options:

  * In the R Console, call `usethis::use_git()`.
  * In RStudio, go to *Tools > Project Options ... > Git/SVN*. Under "Version control system", select "Git". Confirm New Git Repository? Yes!
  * In the shell, with working directory set to the project's directory, do `git init`.

If you used usethis or RStudio to initialize the Git repo, the Project should re-launch in RStudio.
Do that yourself if you did `git init`.
RStudio should now have a Git pane.

## Stage and commit

If your local project was already a Git repo and was up-to-date, move on. Otherwise, you probably need to stage and commit.

* Click the "Git" tab in upper right pane
* Check "Staged" box for all files you want to commit.
  - Default: stage everything
  - When to do otherwise: this will all go to GitHub. So consider if that is
    appropriate for each file. **You can absolutely keep a file locally,
    without committing it to the Git repo and sending to GitHub**. Just let it
    sit there in your Git pane, without being staged. No harm will be done. If
    this is a long-term situation, list the file in `.gitignore`.
* If you're not already in the Git pop-up, click "Commit"
* Type a message in "Commit message".
* Click "Commit"

## Create and connect a GitHub repo

We'll show a few methods for creating a new GitHub repo and connecting it to your local repo.
Pick one.

### Create and connect a GitHub repo with usethis

To use usethis for this task, you must have configured a personal access token (PAT).
This will already by configured for anyone using HTTPS as their protocol, because they are already using the PAT to authenticate for other Git operations.
But if you are an SSH person, you will need to configure a PAT, which is explained in chapter \@ref(https-pat).
It is fine to have both a PAT and SSH keys.

In your project, in the R Console, call:

```{r eval = FALSE}
usethis::use_github()
#> ✓ Creating GitHub repository 'jennybc/myrepo'
#> ✓ Setting remote 'origin' to 'https://github.com/jennybc/myrepo.git'
#> ✓ Pushing 'main' branch to GitHub and setting 'origin/main' as upstream branch
#> ✓ Opening URL 'https://github.com/jennybc/myrepo'
```
    
`usethis::use_github()` does the following:
    
* Creates a new repo on GitHub.
* Configures that new repo as the `origin` remote for the local repo.
* Sets up your local default branch (e.g. `main`) to track same on `origin`.
* Opens the new repo in your browser.

### Create and connect a GitHub repo without usethis

First, you need to create a new repo on GitHub.

```{r echo = FALSE, results = "asis"}
dat <- list(
  repository_name_text = glue::glue("
    Ideally this will be the name of your local project's directory (and \\
    RStudio Project). Why confuse yourself? But it must be a valid \\
    GitHub repo name, which means only letters, digits, `-`, `.`, or `_` are \\
    allowed. For future projects, think about this in advance, i.e. make \\
    sure each project's local name is also a valid GitHub repo name."),
  description_text = glue::glue("
    \"Analysis of the stuff\" or any short description of the project. Write \\
    this for humans."),
  initialize_text = "**DO NOT initialize this repository with anything**."
)
insert <- glue::glue_data(
  dat,
  readr::read_file("child-create-a-github-repo.Rmd"),
  .open = "<<<", .close = ">>>"
)
res <- knitr::knit_child(text = insert, quiet = TRUE)
cat(res, sep = '\n')
```

#### Connect local repo to GitHub repo with RStudio

Click on the "two purple boxes and a white square" in the Git pane.
Click "Add remote".
Paste the GitHub repo's URL here and pick a remote name, almost certainly `origin`.
Now "Add".

We should be back in the "New Branch" dialog (if not, click on the "two purple boxes and a white square" in the Git pane again).
I assume you're on the `main` branch and want it to track `main` on GitHub (or whatever default branch you are using).
Enter `main` as the branch name and make sure "Sync branch with remote" is checked.
Click "Create" (yes, even though the branch already exists).
In the next dialog, choose "overwrite".

#### Connect local repo to GitHub repo with the command line

In a shell, do this, substituting your URL:

```console
git remote add origin https://github.com/jennybc/myrepo.git
```

Push and cement the tracking relationship between your local `main` branch and `main` on GitHub (or whatever your default branch is named):

```console
git push --set-upstream origin main
```

## Confirm the local files propagated to the GitHub remote

Go back to the browser.
I assume we're still viewing your new GitHub repo.

Refresh.

You should see all the project files you committed there.

If this project already had a Git history, it should be reflected on GitHub.

## The end

```{r echo = FALSE, results = "asis"}
cat(readLines("child-the-end-of-repo-setup.Rmd"), sep = '\n')
```
